---
title: "Decoupling of Aboveground and Belowground Intraspecific Trait Variation Along Richness Gradients in Plants"
subtitle: "Public reproduction code: reads the standardized, outlier-removed individual-level dataset and PCA summary directly"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
---

<!--
Fig 1: multidim + PCA-axis ITV | Fig 2: single-trait ITV | Fig 3: Gini coefficient
Fig 4: within/between-species variance decomposition | Fig 5: mean trait shifts
Table S1/S2 | Fig S1: PCA biplots | Fig S2/S3/Table S3: species slopes

This version starts from data/SeedlingTraitData_public.csv and
data/PCA_summary_public.csv instead of the private raw measurement file.
Single-trait columns in the public data are already log10-transformed
(where applicable) and z-standardized, and QC-flagged/dead individuals
are already removed; PCA has already been run and its scores are
included as columns (Above_PC1, Above_PC2, Below_PC1, Below_PC2), with
loadings and variance-explained supplied separately for the Fig. S1
biplot. Every figure and table below reproduces bit-for-bit from these
two files. See prepare_public_data.R for how they were derived from the
private raw dataset.
-->

# Setup

```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
# Fixes the Monte Carlo resampling in Sample_CV() (single-trait ITV, Fig. 2)
# so every render reproduces identical values.
set.seed(42)
```

## Packages

```{r packages}
library(tidyverse)   # data wrangling + ggplot2
library(lme4)        # mixed-effects models
library(lmerTest)    # Satterthwaite P-values for lmer
library(effects)     # allEffects()
library(TPD)          # TPDs(), REND() for multidimensional ITV/FRic
library(ineq)         # Gini coefficient
library(patchwork)   # multi-panel assembly
library(ggpubr)       # get_legend(), as_ggplot()
library(ggthemes)     # theme_foundation()
library(grid)
library(MuMIn)         # r.squaredGLMM()
library(eoffice)       # export to .pptx

# effects loads MASS, which masks dplyr::select()/filter()
select <- dplyr::select
filter <- dplyr::filter
```

## Helper functions

```{r helpers}
# Bao's bias-corrected CV: single-trait ITV metric
CV4 <- function(trait_sample) {
  trait_sample <- trait_sample[!is.na(trait_sample)]
  if (length(trait_sample) > 1) {
    N <- length(trait_sample)
    y_bar <- mean(trait_sample)
    s2_hat <- var(trait_sample)
    cv_2 <- s2_hat / y_bar^2
    cv_1 <- sqrt(cv_2)
    gamma_1 <- sum(((trait_sample - y_bar) / s2_hat^0.5)^3) / N   # skewness term
    gamma_2 <- sum(((trait_sample - y_bar) / s2_hat^0.5)^4) / N   # kurtosis term
    bias2 <- cv_1^3 / N - cv_1 / 4 / N - cv_1^2 * gamma_1 / 2 / N - cv_1 * gamma_2 / 8 / N
    cv4 <- cv_1 - bias2
  } else {
    cv4 <- NA
  }
  return(cv4)
}

# Rarefied ITV: resample to fixed N (999x) for comparability across group sizes
Sample_CV <- function(data, N) {
  if (length(data) <= N) {
    cv4.mean <- CV4(data)
  } else {
    cv <- replicate(999, CV4(sample(data, N)))
    cv4.mean <- mean(cv)
  }
  return(cv4.mean)
}

# Min-max normalization to [0, 1], applied before single-trait ITV
normalize_0_1 <- function(df, columns) {
  for (col_name in columns) {
    min_val <- min(df[[col_name]], na.rm = TRUE)
    max_val <- max(df[[col_name]], na.rm = TRUE)
    df[[col_name]] <- (df[[col_name]] - min_val) / (max_val - min_val)
  }
  return(df)
}

# LRT-based selection of the (1|CRichGroup) composition random effect:
# keep it only if it significantly improves fit over species-only.
select_richness_re <- function(data, response, sp_var = "sp", comp_var = "CRichGroup",
                                richness_term = "log(Richness)", lrt_alpha = 0.05) {
  f_full    <- as.formula(paste(response, "~", richness_term, "+ (1 |", sp_var, ") + (1 |", comp_var, ")"))
  f_reduced <- as.formula(paste(response, "~", richness_term, "+ (1 |", sp_var, ")"))
  # do.call() avoids the lme4 parent.frame() lookup bug inside nested functions
  m_full_ml    <- do.call(lmer, list(formula = f_full, data = data, REML = FALSE))
  m_reduced_ml <- do.call(lmer, list(formula = f_reduced, data = data, REML = FALSE))
  lrt_p <- tryCatch(anova(m_full_ml, m_reduced_ml)[2, "Pr(>Chisq)"], error = function(e) NA_real_)
  use_composition <- !is.na(lrt_p) && lrt_p < lrt_alpha
  re_part <- if (use_composition) paste("(1 |", sp_var, ") + (1 |", comp_var, ")") else paste("(1 |", sp_var, ")")
  list(
    formula = as.formula(paste(response, "~", richness_term, "+", re_part)),
    null_formula = as.formula(paste(response, "~", re_part)),
    used_composition = use_composition,
    lrt_p = lrt_p
  )
}

sp_colors <- c(
  "#5dc1a2", "#13815e", "#e4d79a", "#f54343",
  "#7e0d13", "#1b7c88", "#f7961e", "#5f7385",
  "#ff9999", "#66b3ff")

theme_Publication <- function(base_size = 10, base_family = "sans") {
  (theme_foundation(base_size = base_size, base_family = base_family)
   + theme(plot.title = element_text(face = "bold", size = rel(1.33), hjust = -0.05),
           text = element_text(size = base_size),
           panel.background = element_rect(fill = "white", colour = NA),
           plot.background = element_rect(fill = "white", colour = NA),
           panel.border = element_rect(colour = NA),
           axis.title = element_text(size = base_size + 1),
           axis.title.y = element_text(angle = 90, vjust = 2),
           axis.title.x = element_text(vjust = -0.2),
           axis.text = element_text(size = base_size),
           axis.line = element_line(colour = "black"),
           axis.ticks = element_line(),
           panel.grid.major = element_blank(),
           panel.grid.minor = element_blank(),
           legend.key = element_rect(colour = NA),
           legend.position = "bottom",
           legend.direction = "horizontal",
           legend.key.size = unit(0.2, "cm"),
           legend.margin = ggplot2::margin(0, 0, 0, 0, "mm"),
           legend.title = element_text(size = base_size),
           plot.margin = unit(c(10, 5, 5, 5), "mm"),
           strip.background = element_rect(colour = "#f0f0f0", fill = "#f0f0f0"),
           strip.text = element_text(face = "bold")
   ))
}

# Common ggplot style layer reused by the ITV figures (species-colored points)
common_style <- list(
  theme_Publication(base_size = 10),
  scale_color_manual(values = sp_colors),
  scale_fill_manual(values = sp_colors),
  guides(color = guide_legend(title = "Species", title.position = "left", nrow = 1),
         fill = guide_legend(title = "Species", title.position = "left", nrow = 1)),
  theme(
    legend.position = "bottom",
    legend.direction = "horizontal",
    legend.key.size = unit(0.5, "cm"),
    legend.spacing.x = unit(0.2, "cm"),
    legend.title = element_text(size = 10, face = "bold", hjust = 0.5),
    legend.text = element_text(size = 9, face = "italic"),
    plot.title = element_text(hjust = 0.5, size = 11, face = "bold"),
    axis.title = element_text(size = 11, face = "plain"),
    axis.text = element_text(size = 9, face = "plain", color = "black"),
    axis.text.y = element_text(angle = 90, hjust = 0.5),
    plot.margin = margin(10, 10, 10, 10, "pt")
  )
)

compartment_colors <- c("Aboveground" = "#2F7F6F", "Belowground" = "#9C6B2F")

# Significance threshold for the richness fixed effect (P < 0.1)
SIG_ALPHA <- 0.1

# Shared y-axis label for ITV figures (only shown on the middle-row left panel)
ITV_YLAB <- "Intraspecific trait variability (ITV)"

# Plotmath-parseable "p = 0.028" / "p < 0.001" label (italic p, 3 decimals)
format_p_label <- function(p_value) {
  if (p_value < 0.001) 'paste(italic(p), " < 0.001")' else sprintf('paste(italic(p), " = %.3f")', p_value)
}

# Plotmath-parseable "Name: Slope = 0.067, p = 0.028" label (italic Slope/p)
format_slope_p_label <- function(name, slope, p_value, suffix = "") {
  p_part <- if (p_value < 0.001) '\", \", italic(p), " < 0.001"' else sprintf('\", \", italic(p), " = %.3f"', p_value)
  suffix_part <- if (nzchar(suffix)) sprintf(', "%s"', suffix) else ""
  name_part <- if (nzchar(name)) sprintf('"%s: ", ', name) else ""
  sprintf('paste(%sitalic(Slope), " = %.3f", %s%s)', name_part, slope, p_part, suffix_part)
}

# PCA biplot: scores + trait loading arrows (Fig. S1). Loadings/variance-explained
# come from the precomputed pca_summary table rather than a live principal() fit.
plot_PCA_biplot <- function(point.pca, pca_loadings, x_col, y_col, scale_factor = 2) {
  axis.pca <- pca_loadings %>%
    transmute(Trait, v1 = scale_factor * PC1_loading, v2 = scale_factor * PC2_loading)

  pc1_explained <- round(100 * pca_loadings$PC1_PropVar[1], 1)
  pc2_explained <- round(100 * pca_loadings$PC2_PropVar[1], 1)

  ggplot(point.pca, aes(x = .data[[x_col]], y = .data[[y_col]])) +
    geom_point(aes(color = sp), size = 1.5, alpha = 0.5) +
    xlab(paste0("PC1 (", pc1_explained, "%)")) +
    ylab(paste0("PC2 (", pc2_explained, "%)")) +
    geom_segment(data = axis.pca, aes(x = 0, y = 0, xend = v1, yend = v2),
                 arrow = arrow(length = unit(0.2, "cm")), alpha = 0.9, color = "black", linewidth = 1) +
    geom_text(data = axis.pca, aes(v1, v2, label = Trait), size = 4.5,
              nudge_x = ifelse(axis.pca$v1 > 0, 0.2, -0.2),
              nudge_y = ifelse(axis.pca$v2 > 0, 0.2, -0.2)) +
    theme_Publication() +
    theme(legend.text = element_text(face = "italic"), legend.position = "none") +
    scale_colour_manual(values = sp_colors) +
    scale_fill_manual(values = sp_colors)
}

```

# Data import

```{r data-import}
# Individual-level data: single traits are already log10-transformed (where
# applicable) and z-standardized, QC-flagged measurements and dead/replaced
# individuals are already removed, and PCA axis scores are already attached.
# See prepare_public_data.R (not part of this public pipeline) for how this
# file and the PCA summary below were derived from the private raw dataset.
tda <- read.csv("data/SeedlingTraitData_public.csv", header = TRUE)
tda$sp <- factor(tda$sp, levels = sort(unique(tda$sp)))

pca_summary <- read.csv("data/PCA_summary_public.csv", header = TRUE)

above_traits <- c("LTO", "MLA", "SLA", "LDMC", "SSD", "Chl", "Height")
fineroot_traits <- c("SRL", "RD", "RTD", "SRA")
single_traits <- c(above_traits, fineroot_traits)
pca_traits <- c("Above_PC1", "Above_PC2", "Below_PC1", "Below_PC2")

# Above/below trait tables; NA dropped per organ independently (see Appendix)
above.tda <- na.omit(tda[, colnames(tda) %in% c(
  "Comp", "Richness", "CRichGroup", "Group", "mean_phydist", "pot_number",
  "sp", "ind_ID", "pot_ID", "sp_group", above_traits, "Above_PC1", "Above_PC2")])
below.tda <- na.omit(tda[, colnames(tda) %in% c(
  "Comp", "Richness", "CRichGroup", "Group", "mean_phydist", "pot_number",
  "sp", "ind_ID", "pot_ID", "sp_group", fineroot_traits, "Below_PC1", "Below_PC2")])

# Unique "population" ID = species x richness-composition treatment, the unit
# over which ITV is computed
above.tda$Population <- factor(paste(above.tda$CRichGroup, above.tda$sp),
                                levels = sort(unique(paste(above.tda$CRichGroup, above.tda$sp))))
below.tda$Population <- factor(paste(below.tda$CRichGroup, below.tda$sp),
                                levels = sort(unique(paste(below.tda$CRichGroup, below.tda$sp))))

pop_info_above <- above.tda %>%
  group_by(Population) %>%
  summarise(Richness = mean(Richness, na.rm = TRUE), mean_phydist = mean(mean_phydist, na.rm = TRUE),
            sample_size = n(), sp = unique(sp)[1], CRichGroup = unique(CRichGroup)[1], .groups = "drop")
pop_info_below <- below.tda %>%
  group_by(Population) %>%
  summarise(Richness = mean(Richness, na.rm = TRUE), mean_phydist = mean(mean_phydist, na.rm = TRUE),
            sample_size = n(), sp = unique(sp)[1], CRichGroup = unique(CRichGroup)[1], .groups = "drop")

cat("Aboveground PCA loadings:\n")
print(pca_summary %>% filter(Compartment == "Aboveground") %>% select(-Compartment), row.names = FALSE)
cat("\nBelowground PCA loadings:\n")
print(pca_summary %>% filter(Compartment == "Belowground") %>% select(-Compartment), row.names = FALSE)
```

# Figure 1

```{r fig1-tpd-function}
# TPD/FRic calculator (1-D or 2-D): multidimensional ITV metric
cal_TPD_var <- function(input_tda, pop_info, trait, trait.name) {
  if (!all(trait %in% colnames(input_tda))) stop("Specified traits not found in the dataset")
  kernel.populations <- TPDs(traits = input_tda[, trait, drop = FALSE], species = input_tda$Population)
  functional.indices.populations <- REND(TPDs = kernel.populations)
  population.level <- data.frame(
    trait = trait.name,
    Population = names(functional.indices.populations$species$FRichness),
    Intra_FRic = functional.indices.populations$species$FRichness
  )
  merge(pop_info, population.level, by = "Population")
}
```

## Panel (a-b): 2-D multidimensional ITV (PC1 + PC2 jointly)

```{r fig1-2d}
aboveTPD_FD <- cal_TPD_var(above.tda, pop_info_above, c("Above_PC1", "Above_PC2"), "Aboveground")
belowTPD_FD <- cal_TPD_var(below.tda, pop_info_below, c("Below_PC1", "Below_PC2"), "Belowground")

# Loop at top level (not inside a function) to avoid lme4's parent.frame() bug
multidim_groups <- list(Aboveground = aboveTPD_FD, Belowground = belowTPD_FD)
multidim_plots <- list()
re_selection_log_multidim <- list()

for (group_label in names(multidim_groups)) {
  data_group <- multidim_groups[[group_label]]
  data_group <- data_group[!is.na(data_group$Intra_FRic), ]

  sel <- select_richness_re(data_group, "Intra_FRic")
  model <- do.call(lmer, list(formula = sel$formula, data = data_group, REML = TRUE))
  eff <- allEffects(model)$`log(Richness)`
  effect_data <- data.frame(Richness = eff$x[, 1], value = eff$fit, upper = eff$upper, lower = eff$lower)
  p_value <- summary(model)$coefficients["log(Richness)", "Pr(>|t|)"]
  re_selection_log_multidim[[group_label]] <- data.frame(Panel = group_label, used_CRichGroup = sel$used_composition,
                                                           lrt_p = sel$lrt_p, richness_p = p_value)
  is_significant <- p_value < SIG_ALPHA
  p_label <- format_p_label(p_value)
  y_range <- range(c(data_group$Intra_FRic, effect_data$upper, effect_data$lower), na.rm = TRUE)
  p_y <- y_range[1] + 0.9 * diff(y_range)
  panel_letter <- c(Aboveground = "a", Belowground = "b")[group_label]

  multidim_plots[[group_label]] <- ggplot() +
    geom_jitter(data = data_group, aes(x = log(Richness), y = Intra_FRic, color = sp),
                width = 0.1, shape = 16, size = 2.5, alpha = 0.6) +
    geom_ribbon(data = effect_data, aes(x = log(Richness), ymin = lower, ymax = upper),
                fill = "black", alpha = 0.15, linewidth = 0) +
    geom_line(data = effect_data, aes(x = log(Richness), y = value),
              color = "black", linewidth = 1.2, linetype = ifelse(is_significant, "solid", "dashed")) +
    annotate("text", x = max(log(effect_data$Richness)), y = p_y, label = p_label,
             hjust = 1, vjust = 0, size = 10/.pt, parse = TRUE) +
    common_style + ylab(ITV_YLAB) +
    xlab(if (panel_letter %in% c("a", "b")) NULL else "Richness") +
    ggtitle(paste0("(", panel_letter, ") ", group_label)) +
    theme(plot.title = element_text(hjust = 0, face = "bold")) +
    scale_x_continuous(breaks = log(c(1, 2, 4, 8)), labels = c(1, 2, 4, 8))
}

cat("Random-effects structure selected per panel (Fig. 1a-b):\n")
print(do.call(rbind, re_selection_log_multidim), row.names = FALSE)

# Only the middle row (c) carries the shared y-axis label text; every panel
# keeps its own tick numbers (scales differ by row)
p_multidim_above <- multidim_plots[["Aboveground"]] +
  theme(axis.title.y = element_blank())
p_multidim_below <- multidim_plots[["Belowground"]] +
  theme(axis.title.y = element_blank())
```

## Panel (c-f): 1-D ITV computed separately on PC1 and PC2

```{r fig1-1d}
get_pop_base <- function(df) {
  df %>% group_by(Population) %>%
    summarise(Richness = mean(Richness, na.rm = TRUE), mean_phydist = mean(mean_phydist, na.rm = TRUE),
              sp = unique(sp), CRichGroup = unique(CRichGroup)[1], .groups = "drop")
}
pop_base_above <- get_pop_base(above.tda)
pop_base_below <- get_pop_base(below.tda)

cal_TPD_1D <- function(dataset, base_info, trait_col, label_name) {
  kernel.pop <- TPDs(traits = dataset[, trait_col, drop = FALSE], species = dataset$Population)
  indices <- REND(TPDs = kernel.pop)
  res <- data.frame(Population = names(indices$species$FRichness), Intra_FRic = indices$species$FRichness, Trait = label_name)
  merge(base_info, res, by = "Population")
}

all_results_list <- list(
  "Above_PC1" = cal_TPD_1D(above.tda, pop_base_above, "Above_PC1", "Above_PC1"),
  "Above_PC2" = cal_TPD_1D(above.tda, pop_base_above, "Above_PC2", "Above_PC2"),
  "Below_PC1" = cal_TPD_1D(below.tda, pop_base_below, "Below_PC1", "Below_PC1"),
  "Below_PC2" = cal_TPD_1D(below.tda, pop_base_below, "Below_PC2", "Below_PC2")
)

names_ordered <- c("Above_PC1", "Above_PC2", "Below_PC1", "Below_PC2")
plot_list_fric <- list()
re_selection_log_pcaxis <- list()
for (k in seq_along(names_ordered)) {
  trait_name <- names_ordered[k]
  data.trait <- all_results_list[[trait_name]]
  data.trait <- data.trait[data.trait$Richness > 0 & !is.na(data.trait$Richness), ]

  sel <- select_richness_re(data.trait, "Intra_FRic")
  model.fric <- do.call(lmer, list(formula = sel$formula, data = data.trait, REML = TRUE))
  eff_obj <- allEffects(model.fric)[[1]]
  effect_data <- data.frame(logRichness = log(eff_obj$x[, 1]), fit = eff_obj$fit, upper = eff_obj$upper, lower = eff_obj$lower)
  p_value <- summary(model.fric)$coefficients["log(Richness)", "Pr(>|t|)"]
  re_selection_log_pcaxis[[trait_name]] <- data.frame(Axis = trait_name, used_CRichGroup = sel$used_composition,
                                                       lrt_p = sel$lrt_p, richness_p = p_value)
  is_significant <- p_value < SIG_ALPHA
  p_label <- format_p_label(p_value)
  x_pos <- max(effect_data$logRichness, na.rm = TRUE)
  y_range <- range(c(data.trait$Intra_FRic, effect_data$upper, effect_data$lower), na.rm = TRUE)
  y_pos <- y_range[1] + 0.9 * diff(y_range)
  panel_letter <- c(Above_PC1 = "c", Below_PC1 = "d", Above_PC2 = "e", Below_PC2 = "f")[trait_name]

  p <- ggplot(effect_data, aes(x = logRichness, y = fit)) +
    geom_jitter(data = data.trait, aes(x = log(Richness), y = Intra_FRic, color = sp), width = 0.1, shape = 16, size = 2.5, alpha = 0.6) +
    geom_ribbon(aes(ymin = lower, ymax = upper), fill = "black", alpha = 0.15, linewidth = 0) +
    geom_line(color = "black", linewidth = 1.2, linetype = ifelse(is_significant, "solid", "dashed")) +
    annotate("text", x = x_pos, y = y_pos, label = p_label, hjust = 1, vjust = -0.5, size = 10/.pt, parse = TRUE) +
    common_style + ggtitle(paste0("(", panel_letter, ") ", trait_name)) +
    theme(plot.title = element_text(hjust = 0, face = "bold")) +
    scale_x_continuous(labels = function(x) exp(x), breaks = log(c(1, 2, 4, 8))) +
    scale_y_continuous(expand = expansion(mult = c(0.05, 0.15))) +
    xlab(if (panel_letter %in% c("c", "d")) NULL else "Richness") + ylab(ITV_YLAB)

  # Only panel c (row 2, the middle-left panel) keeps the shared y-axis
  # title; every panel still shows its own tick numbers (scales differ).
  if (panel_letter != "c") {
    p <- p + theme(axis.title.y = element_blank())
  }
  plot_list_fric[[trait_name]] <- p
}

cat("Random-effects structure selected per axis (Fig. 1c-f):\n")
print(do.call(rbind, re_selection_log_pcaxis), row.names = FALSE)
```

## Assemble Fig. 1 (2-D ITV | PC1 ITV | PC2 ITV rows, Above/Below columns)

```{r fig1-assemble, fig.width=9, fig.height=11}
fig1_layout <- "
AB
CD
EF
"

fig1 <- wrap_plots(
  A = p_multidim_above, B = p_multidim_below,
  C = plot_list_fric[["Above_PC1"]], D = plot_list_fric[["Below_PC1"]],
  E = plot_list_fric[["Above_PC2"]], F = plot_list_fric[["Below_PC2"]],
  design = fig1_layout, guides = "collect"
) &
  theme(legend.position = "bottom", plot.margin = margin(3, 4, 3, 4, "pt"))

print(fig1)

ggsave("figures/Figure_1.pdf", plot = fig1, width = 170, height = 170, units = "mm", dpi = 600, device = "pdf")
ggsave("figures/Figure_1.png", plot = fig1, width = 170, height = 170, units = "mm", dpi = 600, device = grDevices::png)
topptx(fig1, filename = "figures/Figure_1.pptx", width = 170 / 25.4, height = 170 / 25.4)
```

# Figure 2

```{r fig2-compute}
# Bao's CV (rarefied to N=20), pooled per species x composition across replicate pots
normalized_data <- normalize_0_1(tda, single_traits)
normalized_data <- na.omit(normalized_data[, c("CRichGroup", "sp", single_traits)])

singleTraitITV <- aggregate(normalized_data[single_traits],
                             by = list(normalized_data$CRichGroup, normalized_data$sp), Sample_CV, 20)
names(singleTraitITV)[1:2] <- c("CRichGroup", "sp")
sp_itv <- singleTraitITV
sp_itv$Richness <- substr(sp_itv$CRichGroup, 5, 5)
sp_itv$Group <- substr(sp_itv$CRichGroup, 8, 8)
sp_itv$Comp <- substr(sp_itv$CRichGroup, 2, 2)
mean_phydist_da <- tda %>% distinct(CRichGroup, mean_phydist) %>% rename(RG = CRichGroup)
sp_itv <- merge(sp_itv, mean_phydist_da, by.x = "CRichGroup", by.y = "RG", all.x = TRUE)
sp_itv <- sp_itv %>% select(sp, Comp, CRichGroup, mean_phydist, Richness, all_of(single_traits))

sp_itv_long1 <- gather(sp_itv, trait, ITV, all_of(single_traits)) %>%
  mutate(Comp = as.factor(Comp),
         trait = factor(trait, levels = single_traits),
         Richness = as.numeric(Richness),
         log_Richness = log(Richness))
```

```{r fig2-model-and-plot, fig.width=9, fig.height=7}
effect_list <- list()
significance_list <- list()
re_selection_log <- list()   # tracks whether (1|CRichGroup) was retained per trait
for (trait in single_traits) {
  data.trait <- sp_itv_long1[sp_itv_long1$trait == trait, ]
  if (nrow(data.trait) == 0) next
  sel <- select_richness_re(data.trait, "ITV")
  model.itv <- do.call(lmer, list(formula = sel$formula, data = data.trait, REML = TRUE))
  eff <- allEffects(model.itv)$`log(Richness)`
  effect_list[[trait]] <- data.frame(logRichness = eff$x, ITV = eff$fit, upper = eff$upper, lower = eff$lower, Trait = trait)
  significance_list[[trait]] <- summary(model.itv)$coefficients["log(Richness)", "Pr(>|t|)"]
  re_selection_log[[trait]] <- data.frame(Trait = trait, used_CRichGroup = sel$used_composition,
                                           lrt_p = sel$lrt_p, richness_p = significance_list[[trait]])
}
effect_data_all <- do.call(rbind, effect_list)

cat("Random-effects structure selected per trait (Fig. 2):\n")
print(do.call(rbind, re_selection_log), row.names = FALSE)

plot_list <- list()
for (trait in single_traits) {
  trait_data <- effect_data_all[effect_data_all$Trait == trait, ]
  raw_data <- sp_itv_long1[sp_itv_long1$trait == trait, ]
  if (nrow(trait_data) == 0 || nrow(raw_data) == 0) next

  p_value <- significance_list[[trait]]
  is_significant <- p_value < SIG_ALPHA
  p_label <- format_p_label(p_value)

  plot_list[[trait]] <- ggplot() +
    geom_jitter(data = raw_data, aes(x = log(Richness), y = ITV, color = sp), width = 0.1, shape = 16, size = 2.5, alpha = 0.6) +
    geom_ribbon(data = trait_data, aes(x = log(Richness), ymin = lower, ymax = upper), fill = "black", alpha = 0.15, linewidth = 0) +
    geom_line(data = trait_data, aes(x = log(Richness), y = ITV), color = "black", linewidth = 1.2,
              linetype = ifelse(is_significant, "solid", "dashed")) +
    annotate("text", x = max(log(trait_data$Richness), na.rm = TRUE),
             y = { yr <- range(c(raw_data$ITV, trait_data$upper, trait_data$lower), na.rm = TRUE); yr[1] + 0.9 * diff(yr) },
             label = p_label, hjust = 1, vjust = 0, size = 10/.pt, parse = TRUE) +
    common_style + xlab("Richness") + ylab(ITV_YLAB) +
    scale_x_continuous(breaks = log(c(1, 2, 4, 8)), labels = c(1, 2, 4, 8))
}

# 11-panel layout: leaf/stem traits (top two rows) then root traits (bottom row)
trait_order <- c("MLA", "SLA", "LDMC", "LTO", "Chl", "Height", "SSD", "SRL", "RD", "RTD", "SRA")
clean_axis_theme <- function(p, letter, label, keep_x = FALSE, keep_y_title = FALSE) {
  p <- p + theme(legend.position = "none") +
    ggtitle(paste0("(", letter, ") ", label)) +
    theme(plot.title = element_text(hjust = 0, face = "bold"))
  if (!keep_x) p <- p + theme(axis.title.x = element_blank(), axis.text.x = element_blank())
  if (!keep_y_title) p <- p + theme(axis.title.y = element_blank())
  p
}
keep_x_traits <- c("SRL", "RD", "RTD", "SRA")
keep_y_title_traits <- c("Chl")                # middle row only
processed_plots <- lapply(seq_along(trait_order), function(i) {
  tr <- trait_order[i]
  clean_axis_theme(plot_list[[tr]], letters[i], tr, keep_x = tr %in% keep_x_traits,
                    keep_y_title = tr %in% keep_y_title_traits)
})
names(processed_plots) <- trait_order

fig2_layout <- "
ABCD
EFG#
HIJK
"
fig2 <- wrap_plots(processed_plots[trait_order], design = fig2_layout, guides = "collect") &
  theme(legend.position = "bottom", plot.margin = margin(3, 4, 3, 4, "pt"))

print(fig2)

ggsave("figures/Figure_2.pdf", plot = fig2, width = 170, height = 142, units = "mm", dpi = 600, device = "pdf")
ggsave("figures/Figure_2.png", plot = fig2, width = 170, height = 142, units = "mm", dpi = 600, device = grDevices::png)
topptx(fig2, filename = "figures/Figure_2.pptx", width = 170 / 25.4, height = 142 / 25.4)
```

# Figure 3

```{r fig3a-gini-compute}
# Gini coefficient (bias-corrected, Weiner & Solbrig 1984): size inequality per pot
clean_data_gini <- tda %>%
  filter(Comp %in% c("TRUE", "FALSE")) %>%
  filter(!is.na(ABM) & ABM > 0 & !is.na(UBM) & UBM > 0 & Richness > 0)

pot_stats <- clean_data_gini %>%
  group_by(CRichGroup, Richness, pot_ID) %>%
  summarise(n_ind = n(),
            gini_raw_ABM = ineq(ABM, type = "Gini"),
            gini_raw_UBM = ineq(UBM, type = "Gini"), .groups = "drop") %>%
  filter(n_ind > 1) %>%
  mutate(Gini_ABM = gini_raw_ABM * (n_ind / (n_ind - 1)),   # small-sample correction (Eq. 2)
         Gini_UBM = gini_raw_UBM * (n_ind / (n_ind - 1)),
         log_Richness = log(Richness))

model_abm <- lmer(Gini_ABM ~ log_Richness + (1 | CRichGroup), data = pot_stats)
model_ubm <- lmer(Gini_UBM ~ log_Richness + (1 | CRichGroup), data = pot_stats)

get_stats_text <- function(model, name, suffix = "") {
  coefs <- summary(model)$coefficients
  slope <- coefs["log_Richness", "Estimate"]
  p_val <- coefs["log_Richness", "Pr(>|t|)"]
  format_slope_p_label(name, slope, p_val, suffix)
}
label_text_abm <- get_stats_text(model_abm, "Aboveground")
label_text_ubm <- get_stats_text(model_ubm, "Belowground")

# Paired t-test per richness level: is aboveground Gini > belowground Gini?
t_test_stats <- pot_stats %>%
  group_by(Richness) %>%
  summarise(p_val = t.test(Gini_ABM, Gini_UBM, paired = TRUE)$p.value,
            y_pos_raw = max(c(Gini_ABM, Gini_UBM), na.rm = TRUE), .groups = "drop") %>%
  mutate(log_Richness = log(Richness),
         significance = case_when(p_val < 0.001 ~ "***", p_val < 0.01 ~ "**", p_val < 0.05 ~ "*", TRUE ~ "ns"),
         y_pos = pmin(y_pos_raw * 1.05, 1.03))
```

```{r fig3-gini-plot, fig.width=8, fig.height=7}
eff_abm <- as.data.frame(effect("log_Richness", model_abm))
eff_ubm <- as.data.frame(effect("log_Richness", model_ubm))
eff_combined <- rbind(mutate(eff_abm, Type = "Aboveground"), mutate(eff_ubm, Type = "Belowground"))
eff_combined$Type <- factor(eff_combined$Type, levels = c("Aboveground", "Belowground"))

plot_data_long <- pot_stats %>%
  select(Richness, Gini_ABM, Gini_UBM) %>%
  pivot_longer(cols = c(Gini_ABM, Gini_UBM), names_to = "Type", values_to = "Gini") %>%
  mutate(Type = recode(Type, "Gini_ABM" = "Aboveground", "Gini_UBM" = "Belowground"),
         log_Richness = log(Richness))
plot_data_long$Type <- factor(plot_data_long$Type, levels = c("Aboveground", "Belowground"))

ann_text <- data.frame(log_Richness = log(1), Gini = c(1.18, 1.1),
                        label = c(label_text_abm, label_text_ubm),
                        Type = factor(c("Aboveground", "Belowground"), levels = c("Aboveground", "Belowground")))

dodge_width <- 0.2

# Shared panel builder: boxplot + jitter (+ optional fitted line/ribbon) + P annotation
build_gini_style_panel <- function(plot_data_long, eff_combined, ann_text, t_test_stats,
                                    y_label, x_breaks, x_labels, y_limits = NULL, y_breaks = NULL, tag = NULL, hide_x = FALSE,
                                    show_line = TRUE) {
  p <- ggplot() +
    geom_jitter(data = plot_data_long, aes(x = log_Richness, y = Value, color = Type),
                position = position_jitterdodge(jitter.width = 0.1, dodge.width = dodge_width), size = 2, alpha = 0.4, shape = 16) +
    geom_boxplot(data = plot_data_long, aes(x = log_Richness, y = Value, group = interaction(Richness, Type), color = Type),
                 position = position_dodge(width = dodge_width), width = 0.2, alpha = 0, outlier.shape = NA, linewidth = 0.8) +
    { if (show_line) list(
        geom_ribbon(data = eff_combined, aes(x = log_Richness, ymin = lower, ymax = upper, fill = Type), alpha = 0.15, show.legend = FALSE),
        geom_line(data = eff_combined, aes(x = log_Richness, y = fit, color = Type), linewidth = 1.2)
      ) } +
    geom_text(data = ann_text, aes(x = log_Richness, y = Value, label = label, color = Type),
              hjust = 0, vjust = 0.5, size = 10/.pt, show.legend = FALSE, parse = TRUE) +
    geom_text(data = t_test_stats, aes(x = log_Richness, y = y_pos, label = significance), size = 10/.pt, color = "black") +
    theme_Publication(base_size = 10) +
    scale_color_manual(values = compartment_colors, name = "Compartment") +
    scale_fill_manual(values = compartment_colors, name = "Compartment") +
    theme(legend.position = "bottom", plot.tag = element_text(size = 11, face = "bold", family = "sans"),
          plot.margin = margin(3, 4, 3, 4, "pt"),
          legend.title = element_text(size = 10, face = "bold"),
          legend.text = element_text(size = 9, face = "plain"),
          legend.key.size = unit(0.5, "cm"),
          axis.title = element_text(size = 11, face = "plain"),
          axis.text = element_text(size = 9, face = "plain", color = "black")) +
    labs(y = y_label, x = if (hide_x) NULL else "Richness", tag = if (!is.null(tag)) paste0("(", tag, ")") else NULL) +
    scale_x_continuous(breaks = x_breaks, labels = x_labels)
  if (!is.null(y_limits)) {
    p <- p + scale_y_continuous(limits = y_limits, breaks = if (!is.null(y_breaks)) y_breaks else seq(y_limits[1], y_limits[2], 0.2), expand = expansion(mult = c(0.05, 0)))
  } else {
    p <- p + scale_y_continuous(expand = expansion(mult = c(0.05, 0.18)))
  }
  p
}

# Gini coefficient panel (Fig. 3, single panel)
ann_text_a <- ann_text %>% rename(Value = Gini)
p3a <- build_gini_style_panel(rename(plot_data_long, Value = Gini), eff_combined, ann_text_a, t_test_stats,
                               "Gini coefficient of biomass inequality", log(c(1, 2, 4, 8)), c("1", "2", "4", "8"),
                               y_limits = c(0, 1.25), y_breaks = seq(0, 1, 0.2), hide_x = FALSE)

print(p3a)
ggsave("figures/Figure_3.pdf", plot = p3a, width = 120, height = 100, units = "mm", dpi = 600, device = "pdf")
ggsave("figures/Figure_3.png", plot = p3a, width = 120, height = 100, units = "mm", dpi = 600, device = grDevices::png)
topptx(p3a, filename = "figures/Figure_3.pptx", width = 120 / 25.4, height = 100 / 25.4)
```

# Figure 4

```{r fig4-compute}
# Within/between-species biomass SS decomposition, per pot
clean_data <- tda %>%
  filter(Comp %in% c("TRUE", "FALSE")) %>%
  filter(!is.na(ABM) & ABM > 0 & !is.na(UBM) & UBM > 0 & Richness > 0) %>%
  mutate(log_ABM = log(ABM), log_UBM = log(UBM))

sp_level <- clean_data %>%
  group_by(pot_ID, Richness, CRichGroup, sp) %>%
  summarise(n_sp = n(),
            sp_mean_ABM = mean(log_ABM, na.rm = TRUE), sp_mean_UBM = mean(log_UBM, na.rm = TRUE),
            SSW_sp_ABM = sum((log_ABM - sp_mean_ABM)^2, na.rm = TRUE),
            SSW_sp_UBM = sum((log_UBM - sp_mean_UBM)^2, na.rm = TRUE), .groups = "drop")

pot_level <- sp_level %>%
  group_by(pot_ID, Richness, CRichGroup) %>%
  mutate(n_pot = sum(n_sp), n_sp_pot = n(),
         grand_mean_ABM = sum(n_sp * sp_mean_ABM) / n_pot,
         grand_mean_UBM = sum(n_sp * sp_mean_UBM) / n_pot) %>%
  summarise(df_within = max(n_pot - n_sp_pot), df_between = max(n_sp_pot - 1),
            SSW_ABM = sum(SSW_sp_ABM, na.rm = TRUE), SSW_UBM = sum(SSW_sp_UBM, na.rm = TRUE),
            SSB_ABM = sum(n_sp * (sp_mean_ABM - grand_mean_ABM)^2, na.rm = TRUE),
            SSB_UBM = sum(n_sp * (sp_mean_UBM - grand_mean_UBM)^2, na.rm = TRUE), .groups = "drop") %>%
  mutate(PropB_ABM = ifelse((SSW_ABM + SSB_ABM) > 0, SSB_ABM / (SSW_ABM + SSB_ABM), NA),
         PropB_UBM = ifelse((SSW_UBM + SSB_UBM) > 0, SSB_UBM / (SSW_UBM + SSB_UBM), NA),
         log_Richness = log(Richness))

# Panel (b/c): within-species SS, richness = 1, 2, 4, split by compartment
intra_wide <- pot_level %>% filter(Richness %in% c(1, 2, 4))
mod_intra_abm <- lmer(SSW_ABM ~ log_Richness + (1 | CRichGroup), data = intra_wide)
mod_intra_ubm <- lmer(SSW_UBM ~ log_Richness + (1 | CRichGroup), data = intra_wide)
eff_intra_abm <- as.data.frame(effect("log_Richness", mod_intra_abm))
eff_intra_ubm <- as.data.frame(effect("log_Richness", mod_intra_ubm))

# Panel (e/f): between-species variance proportion, richness = 2, 4, 8, split
# by compartment. NB: at richness=8 each pot has 1 individual/species, so
# within-pot SSW=0 and PropB is forced to 1 for most such pots (ceiling,
# not pure biology).
inter_wide_plot <- pot_level %>% filter(Richness %in% c(2, 4, 8))
mod_inter_abm <- lmer(PropB_ABM ~ log_Richness + (1 | CRichGroup), data = inter_wide_plot)
mod_inter_ubm <- lmer(PropB_UBM ~ log_Richness + (1 | CRichGroup), data = inter_wide_plot)
eff_inter_abm <- as.data.frame(effect("log_Richness", mod_inter_abm))
eff_inter_ubm <- as.data.frame(effect("log_Richness", mod_inter_ubm))
```

```{r fig4-plot, fig.width=9, fig.height=8}
# Single-compartment panel: boxplot + jitter + fitted line/ribbon + Slope/p
build_single_panel <- function(data, value_col, model, eff_data, color, y_label, x_breaks, x_labels,
                                y_limits = NULL, y_breaks = NULL, tag, title, hide_x = FALSE, show_y = TRUE) {
  label <- get_stats_text(model, "")
  y_range <- range(c(data[[value_col]], eff_data$upper, eff_data$lower), na.rm = TRUE)
  p_y <- y_range[1] + 0.95 * diff(y_range)

  ggplot() +
    geom_jitter(data = data, aes(x = log(Richness), y = .data[[value_col]]),
                width = 0.1, size = 2, alpha = 0.4, shape = 16, color = color) +
    geom_boxplot(data = data, aes(x = log(Richness), y = .data[[value_col]], group = Richness),
                 color = color, width = 0.2, alpha = 0, outlier.shape = NA, linewidth = 0.8) +
    geom_ribbon(data = eff_data, aes(x = log_Richness, ymin = lower, ymax = upper), fill = color, alpha = 0.15) +
    geom_line(data = eff_data, aes(x = log_Richness, y = fit), color = color, linewidth = 1.2) +
    annotate("text", x = min(log(data$Richness)), y = p_y, label = label,
             hjust = 0, vjust = 1, size = 10/.pt, parse = TRUE, color = color) +
    theme_Publication(base_size = 10) +
    theme(axis.title = element_text(size = 11, face = "plain"),
          axis.text = element_text(size = 9, face = "plain", color = "black"),
          legend.position = "none",
          plot.title = element_text(hjust = 0, face = "bold", size = 11),
          plot.margin = margin(3, 4, 3, 4, "pt")) +
    ggtitle(paste0("(", tag, ") ", title)) +
    xlab(if (hide_x) NULL else "Richness") +
    ylab(if (show_y) y_label else NULL) +
    scale_x_continuous(breaks = x_breaks, labels = x_labels) +
    { if (!is.null(y_limits)) scale_y_continuous(limits = y_limits, breaks = y_breaks, expand = expansion(mult = c(0.05, 0))) }
}

p4a <- build_single_panel(intra_wide, "SSW_ABM", mod_intra_abm, eff_intra_abm, unname(compartment_colors["Aboveground"]),
                           "Within-species SS (log-biomass)", log(c(1, 2, 4)), c("1", "2", "4"),
                           tag = "a", title = "Aboveground", hide_x = TRUE)
p4b <- build_single_panel(intra_wide, "SSW_UBM", mod_intra_ubm, eff_intra_ubm, unname(compartment_colors["Belowground"]),
                           "Within-species SS (log-biomass)", log(c(1, 2, 4)), c("1", "2", "4"),
                           tag = "b", title = "Belowground", hide_x = TRUE, show_y = FALSE)
p4c <- build_single_panel(inter_wide_plot, "PropB_ABM", mod_inter_abm, eff_inter_abm, unname(compartment_colors["Aboveground"]),
                           "Proportion of variance between species", log(c(2, 4, 8)), c("2", "4", "8"),
                           y_limits = c(0, 1.3), y_breaks = c(0, 0.25, 0.5, 0.75, 1), tag = "c", title = "Aboveground")
p4d <- build_single_panel(inter_wide_plot, "PropB_UBM", mod_inter_ubm, eff_inter_ubm, unname(compartment_colors["Belowground"]),
                           "Proportion of variance between species", log(c(2, 4, 8)), c("2", "4", "8"),
                           y_limits = c(0, 1.3), y_breaks = c(0, 0.25, 0.5, 0.75, 1), tag = "d", title = "Belowground", show_y = FALSE)

fig4_decomp <- (p4a | p4b) / (p4c | p4d) &
  theme(plot.margin = margin(3, 4, 3, 4, "pt"))

print(fig4_decomp)

ggsave("figures/Figure_4.pdf", plot = fig4_decomp, width = 170, height = 160, units = "mm", dpi = 600, device = "pdf")
ggsave("figures/Figure_4.png", plot = fig4_decomp, width = 170, height = 160, units = "mm", dpi = 600, device = grDevices::png)
topptx(fig4_decomp, filename = "figures/Figure_4.pptx", width = 170 / 25.4, height = 160 / 25.4)
```

# Figure 5

```{r fig5-compute}
# mean trait value ~ log(Richness) + (1|sp), per trait + PCA axis
meta_cols <- c("ind_ID", "Richness", "CRichGroup", "sp", "mean_phydist")

data_mean <- tda %>%
  select(all_of(meta_cols), all_of(single_traits)) %>%
  left_join(above.tda %>% select(ind_ID, Above_PC1, Above_PC2), by = "ind_ID") %>%
  left_join(below.tda %>% select(ind_ID, Below_PC1, Below_PC2), by = "ind_ID")

all_mean_traits <- c(single_traits, "Above_PC1", "Above_PC2", "Below_PC1", "Below_PC2")

sp_mean_by_richness_group <- data_mean %>%
  group_by(Richness, CRichGroup, sp) %>%
  summarise(across(all_of(all_mean_traits), ~ mean(., na.rm = TRUE)), .groups = "drop") %>%
  mutate(Richness = as.numeric(Richness)) %>%
  filter(Richness != 0) %>%
  mutate(log_Richness = log(Richness))

traits_ordered <- c("LTO", "MLA", "SLA", "LDMC", "SSD", "Chl", "Height",
                     "Above_PC1", "Above_PC2", "SRL", "RD", "RTD", "SRA", "Below_PC1", "Below_PC2")

results.mean_trait <- data.frame(Trait = traits_ordered, p.value = NA, estimate = NA, error = NA)
results.mean_trait$Trait <- factor(results.mean_trait$Trait, levels = rev(traits_ordered))
re_selection_log_fig5 <- list()

for (trait in traits_ordered) {
  data.trait <- sp_mean_by_richness_group %>%
    select(Richness, log_Richness, sp, CRichGroup, all_of(trait)) %>%
    rename(mean_trait_value = trait)

  sel <- select_richness_re(data.trait, "mean_trait_value", richness_term = "log_Richness")

  model.mean_trait <- do.call(lmer, list(formula = sel$formula, data = data.trait, REML = TRUE))
  coef_summary <- summary(model.mean_trait)$coefficients
  # Wald t-test P-value for log(Richness): used both for the significance
  # fill in the plot and for Table S1, so the two stay consistent.
  wald_p <- coef_summary["log_Richness", "Pr(>|t|)"]
  results.mean_trait[results.mean_trait$Trait == trait, "p.value"] <- wald_p
  results.mean_trait[results.mean_trait$Trait == trait, c("estimate", "error")] <-
    coef_summary["log_Richness", c("Estimate", "Std. Error")]

  re_selection_log_fig5[[trait]] <- data.frame(Trait = trait, used_CRichGroup = sel$used_composition,
                                                lrt_p = sel$lrt_p, richness_p = wald_p)
}

cat("Random-effects structure selected per trait (Fig. 5):\n")
print(do.call(rbind, re_selection_log_fig5), row.names = FALSE)

results.mean_trait$sig <- with(results.mean_trait,
  factor(ifelse(p.value < 0.05, "p < 0.05", ifelse(p.value < 0.1, "p < 0.10 .", "ns")),
         levels = c("p < 0.05", "p < 0.10 .", "ns")))

belowground_traits <- c("SRL", "RD", "RTD", "SRA", "Below_PC1", "Below_PC2")
results.mean_trait$Compartment <- factor(ifelse(results.mean_trait$Trait %in% belowground_traits, "Belowground", "Aboveground"),
                                          levels = c("Aboveground", "Belowground"))
write.csv(results.mean_trait, "tables/mean_trait_analysis_results.csv", row.names = FALSE)
```

```{r fig5-plot, fig.width=6, fig.height=5}
fig5 <- ggplot(results.mean_trait, aes(x = estimate, y = Trait, color = Compartment, fill = sig)) +
  geom_vline(xintercept = 0, linetype = 2, color = "grey60", linewidth = 1) +
  geom_errorbar(aes(xmin = estimate - 1.96 * error, xmax = estimate + 1.96 * error), width = 0.2, linewidth = 0.8) +
  geom_point(shape = 21, stroke = 1.2, size = 3) +
  theme_Publication(base_size = 10) +
  theme(axis.title = element_text(size = 11, face = "plain"),
        axis.text = element_text(size = 9, face = "plain", color = "black"),
        legend.position = "bottom",
        legend.title = element_text(size = 10, face = "bold"),
        legend.text = element_text(size = 9, face = "plain"),
        legend.key.size = unit(0.5, "cm"),
        plot.margin = margin(3, 4, 3, 4, "pt")) +
  scale_fill_manual(values = c("p < 0.05" = "black", "p < 0.10 ." = "grey80", "ns" = "white"), guide = "none") +
  scale_color_manual(values = compartment_colors, name = "Compartment") +
  ylab("Trait") + xlab("Regression coefficients (Richness)")

print(fig5)
ggsave("figures/Figure_5.pdf", plot = fig5, width = 120, height = 100, units = "mm", dpi = 600, device = "pdf")
ggsave("figures/Figure_5.png", plot = fig5, width = 120, height = 100, units = "mm", dpi = 600, device = grDevices::png)
topptx(fig5, filename = "figures/Figure_5.pptx", width = 120 / 25.4, height = 100 / 25.4)
```

# Table S1

```{r tableS1}
# Composition random-effect (LRT) decision per metric, Fig. 1 and Fig. 2
above_traits_set <- above_traits
tableS1_multidim <- do.call(rbind, re_selection_log_multidim) %>%
  rename(Metric = Panel) %>%
  mutate(Compartment = Metric, Type = "Multidimensional (2-D) ITV")

tableS1_pcaxis <- do.call(rbind, re_selection_log_pcaxis) %>%
  rename(Metric = Axis) %>%
  mutate(Compartment = ifelse(grepl("^Above", Metric), "Aboveground", "Belowground"),
         Type = "PCA-axis (1-D) ITV")

tableS1_singletrait <- do.call(rbind, re_selection_log) %>%
  rename(Metric = Trait) %>%
  mutate(Compartment = ifelse(Metric %in% above_traits_set, "Aboveground", "Belowground"),
         Type = "Single-trait ITV")

tableS1_meantrait <- do.call(rbind, re_selection_log_fig5) %>%
  rename(Metric = Trait) %>%
  mutate(Compartment = ifelse(Metric %in% belowground_traits, "Belowground", "Aboveground"),
         Type = "Species-level mean trait value")

tableS1 <- bind_rows(tableS1_multidim, tableS1_pcaxis, tableS1_singletrait, tableS1_meantrait) %>%
  mutate(Composition_included = ifelse(used_CRichGroup, "Yes", "No")) %>%
  select(Type, Compartment, Metric, LRT_P = lrt_p, Composition_included, Richness_effect_P = richness_p) %>%
  mutate(across(c(LRT_P, Richness_effect_P), ~ round(., 4)))

print(tableS1, row.names = FALSE)
write.csv(tableS1, "tables/TableS1_composition_LRT.csv", row.names = FALSE)

# Publication-formatted version: fixed 3 decimals, "<0.001" for very small
# P-values, and a non-redundant Metric label for the two multidimensional
# ITV rows (their Metric otherwise just repeats the Compartment column).
format_p <- function(p) ifelse(p < 0.001, "<0.001", sprintf("%.3f", p))
tableS1_manuscript <- tableS1 %>%
  mutate(Metric = ifelse(Type == "Multidimensional (2-D) ITV", "TPD FRic (all traits)", Metric),
         LRT_P = format_p(LRT_P),
         Richness_effect_P = format_p(Richness_effect_P)) %>%
  rename(`ITV type` = Type, `LRT P` = LRT_P, `Composition retained` = Composition_included,
         `Richness effect P` = Richness_effect_P)
write.csv(tableS1_manuscript, "tables/TableS1_manuscript.csv", row.names = FALSE)
```

# Table S2

```{r tableS2}
# log(Richness) x Compartment interaction, pooled across all ITV metrics (z-standardized)
df_single <- sp_itv_long1 %>%
  filter(trait %in% single_traits) %>%
  mutate(Compartment = ifelse(trait %in% above_traits, "Aboveground", "Belowground"),
         Measure_ID = trait, Value = ITV, Data_Type = "Single_Trait") %>%
  select(sp, Richness, CRichGroup, Compartment, Measure_ID, Value, Data_Type)

df_pca <- do.call(rbind, lapply(names(all_results_list), function(n) {
  d <- all_results_list[[n]]; d$Measure_ID <- n; d
})) %>%
  mutate(Compartment = ifelse(grepl("Above", Measure_ID), "Aboveground", "Belowground"),
         Value = Intra_FRic, Data_Type = "PCA_Axis") %>%
  select(sp, Richness, CRichGroup, Compartment, Measure_ID, Value, Data_Type)

df_multi <- rbind(
  aboveTPD_FD %>% mutate(Compartment = "Aboveground"),
  belowTPD_FD %>% mutate(Compartment = "Belowground")
) %>%
  mutate(Measure_ID = "Multi_FRic", Value = Intra_FRic, Data_Type = "Multi_Trait") %>%
  select(sp, Richness, CRichGroup, Compartment, Measure_ID, Value, Data_Type)

df_all_pooled <- rbind(df_single, df_pca, df_multi) %>%
  group_by(Measure_ID) %>%
  mutate(Standardized_Value = scale(Value)) %>%
  ungroup()

sel_pooled <- select_richness_re(df_all_pooled, "Standardized_Value", richness_term = "log(Richness) * Compartment")
cat("Random-effects structure selected for Table S2 pooled model: used_CRichGroup =",
    sel_pooled$used_composition, ", LRT P =", signif(sel_pooled$lrt_p, 3), "\n")

model_all_pooled <- do.call(lmer, list(formula = sel_pooled$formula, data = df_all_pooled, REML = TRUE))

anova(model_all_pooled)
r.squaredGLMM(model_all_pooled)

# Per-predictor coefficient table (Estimate/SE/df/t/P), exported so manuscript
# Table S2 is transcribed directly from this file rather than by hand.
tableS2_coefs <- as.data.frame(summary(model_all_pooled)$coefficients)
tableS2_coefs <- cbind(Predictor = rownames(tableS2_coefs), tableS2_coefs)
rownames(tableS2_coefs) <- NULL
names(tableS2_coefs) <- c("Predictor", "Estimate", "Std_Error", "df", "t_value", "P_value")
print(tableS2_coefs, row.names = FALSE)
write.csv(tableS2_coefs, "tables/TableS2_coefficients.csv", row.names = FALSE)
```

# Figure S1

```{r figS1, fig.width=10, fig.height=5.5}
above_pca_plot <- plot_PCA_biplot(above.tda, pca_summary %>% filter(Compartment == "Aboveground"), "Above_PC1", "Above_PC2", scale_factor = 2) +
  ggtitle("(a) Aboveground") + common_style + theme(plot.title = element_text(hjust = 0, face = "bold"))
froot_pca_plot <- plot_PCA_biplot(below.tda, pca_summary %>% filter(Compartment == "Belowground"), "Below_PC1", "Below_PC2", scale_factor = 3) +
  ggtitle("(b) Belowground") + common_style + theme(plot.title = element_text(hjust = 0, face = "bold"))

figS1 <- above_pca_plot + froot_pca_plot +
  plot_layout(ncol = 2, guides = "collect") &
  theme(legend.position = "bottom", legend.box.margin = margin(t = -10, b = 0),
        plot.margin = margin(3, 4, 3, 4, "pt"))

print(figS1)
ggsave("figures/Figure_S1.pdf", plot = figS1, width = 170, height = 91, units = "mm", dpi = 600, device = "pdf")
ggsave("figures/Figure_S1.png", plot = figS1, width = 170, height = 91, units = "mm", dpi = 600, device = grDevices::png)
topptx(figS1, filename = "figures/Figure_S1.pptx", width = 170 / 25.4, height = 91 / 25.4)
```

# Figure S2

```{r figS-species-helpers}
# Per-species lm(response ~ log(Richness)), independent of variance-component estimation
fit_species_lines <- function(data, response_col, panel_label) {
  data <- data[!is.na(data[[response_col]]), ]
  species_list <- sort(unique(data$sp))
  lines <- list()
  rows <- list()
  for (s in species_list) {
    d <- data[data$sp == s, ]
    if (nrow(d) < 3 || length(unique(d$Richness)) < 2) next
    f <- as.formula(paste(response_col, "~ log(Richness)"))
    m <- lm(f, data = d)
    sm <- summary(m)
    slope <- sm$coefficients[2, 1]
    slope_se <- sm$coefficients[2, 2]
    p_value <- sm$coefficients[2, 4]
    r2 <- sm$r.squared

    newx <- data.frame(Richness = seq(min(d$Richness), max(d$Richness), length.out = 50))
    pred <- predict(m, newdata = newx)
    lines[[s]] <- data.frame(sp = s, logRichness = log(newx$Richness), fit = pred,
                              is_significant = p_value < SIG_ALPHA)
    rows[[s]] <- data.frame(Panel = panel_label, sp = s, N = nrow(d), slope = slope, slope_se = slope_se,
                             r_squared = r2, p_value = p_value)
  }
  list(lines = if (length(lines)) do.call(rbind, lines) else NULL,
       points = data.frame(sp = d$sp[0], Richness = numeric(0), value = numeric(0)),
       table = if (length(rows)) do.call(rbind, rows) else NULL) -> out
  out$points <- data.frame(sp = data$sp, Richness = data$Richness, value = data[[response_col]])
  out
}

plot_species_overlay <- function(fit_result, title, letter, hide_x = FALSE, hide_y = FALSE, show_y_title = TRUE) {
  p <- ggplot() +
    geom_jitter(data = fit_result$points, aes(x = log(Richness), y = value, color = sp),
                width = 0.05, shape = 16, size = 1.8, alpha = 0.5) +
    geom_line(data = fit_result$lines, aes(x = logRichness, y = fit, color = sp, linetype = is_significant),
              linewidth = 0.9) +
    scale_linetype_manual(values = c(`TRUE` = "solid", `FALSE` = "dashed"), guide = "none") +
    common_style + ggtitle(paste0("(", letter, ") ", title)) +
    theme(plot.title = element_text(hjust = 0, face = "bold")) +
    xlab("Richness") + ylab(ITV_YLAB) +
    scale_x_continuous(breaks = log(c(1, 2, 4, 8)), labels = c(1, 2, 4, 8))
  if (hide_x) p <- p + theme(axis.title.x = element_blank(), axis.text.x = element_blank())
  if (hide_y) {
    p <- p + theme(axis.title.y = element_blank(), axis.text.y = element_blank())
  } else if (!show_y_title) {
    p <- p + theme(axis.title.y = element_blank())
  }
  p
}
```

```{r figS2-compute}
fit_above_multidim <- fit_species_lines(aboveTPD_FD, "Intra_FRic", "Aboveground")
fit_below_multidim <- fit_species_lines(belowTPD_FD, "Intra_FRic", "Belowground")
fit_above_pc1 <- fit_species_lines(all_results_list[["Above_PC1"]], "Intra_FRic", "Above_PC1")
fit_below_pc1 <- fit_species_lines(all_results_list[["Below_PC1"]], "Intra_FRic", "Below_PC1")
fit_above_pc2 <- fit_species_lines(all_results_list[["Above_PC2"]], "Intra_FRic", "Above_PC2")
fit_below_pc2 <- fit_species_lines(all_results_list[["Below_PC2"]], "Intra_FRic", "Below_PC2")

species_slopes_S2 <- rbind(fit_above_multidim$table, fit_below_multidim$table,
                            fit_above_pc1$table, fit_below_pc1$table,
                            fit_above_pc2$table, fit_below_pc2$table)
cat("Per-species richness-ITV fits (multidimensional + PCA-axis FRic):\n")
print(species_slopes_S2, row.names = FALSE)
write.csv(species_slopes_S2, "tables/TableS3_species_slopes_multidim_pca.csv", row.names = FALSE)
```

```{r figS2, fig.width=9, fig.height=11}
figS2 <- wrap_plots(
  A = plot_species_overlay(fit_above_multidim, "Aboveground", "a", hide_x = TRUE, show_y_title = FALSE),
  B = plot_species_overlay(fit_below_multidim, "Belowground", "b", hide_x = TRUE, show_y_title = FALSE),
  C = plot_species_overlay(fit_above_pc1, "Above_PC1", "c", hide_x = TRUE),
  D = plot_species_overlay(fit_below_pc1, "Below_PC1", "d", hide_x = TRUE, show_y_title = FALSE),
  E = plot_species_overlay(fit_above_pc2, "Above_PC2", "e", show_y_title = FALSE),
  F = plot_species_overlay(fit_below_pc2, "Below_PC2", "f", show_y_title = FALSE),
  design = "AB\nCD\nEF", guides = "collect"
) &
  theme(legend.position = "bottom", plot.margin = margin(3, 4, 3, 4, "pt"))

print(figS2)
ggsave("figures/Figure_S2.pdf", plot = figS2, width = 170, height = 170, units = "mm", dpi = 600, device = "pdf")
ggsave("figures/Figure_S2.png", plot = figS2, width = 170, height = 170, units = "mm", dpi = 600, device = grDevices::png)
topptx(figS2, filename = "figures/Figure_S2.pptx", width = 170 / 25.4, height = 170 / 25.4)
```

# Figure S3

```{r figS3-compute}
single_trait_fits <- lapply(trait_order, function(tr) {
  fit_species_lines(sp_itv_long1[sp_itv_long1$trait == tr, ], "ITV", tr)
})
names(single_trait_fits) <- trait_order

species_slopes_S3 <- do.call(rbind, lapply(single_trait_fits, `[[`, "table"))
cat("Per-species richness-ITV fits (single-trait ITV):\n")
print(species_slopes_S3, row.names = FALSE)
write.csv(species_slopes_S3, "tables/TableS3_species_slopes_singletrait.csv", row.names = FALSE)
```

```{r figS3, fig.width=9, fig.height=7}
keep_x_traits <- c("SRL", "RD", "RTD", "SRA")
plots_S3 <- lapply(seq_along(trait_order), function(i) {
  tr <- trait_order[i]
  plot_species_overlay(single_trait_fits[[tr]], tr, letters[i],
                        hide_x = !(tr %in% keep_x_traits),
                        show_y_title = tr %in% keep_y_title_traits)
})
names(plots_S3) <- trait_order

figS3_layout <- "
ABCD
EFG#
HIJK
"
figS3 <- wrap_plots(plots_S3[trait_order], design = figS3_layout, guides = "collect") &
  theme(legend.position = "bottom", plot.margin = margin(3, 4, 3, 4, "pt"))

print(figS3)
ggsave("figures/Figure_S3.pdf", plot = figS3, width = 170, height = 142, units = "mm", dpi = 600, device = "pdf")
ggsave("figures/Figure_S3.png", plot = figS3, width = 170, height = 142, units = "mm", dpi = 600, device = grDevices::png)
topptx(figS3, filename = "figures/Figure_S3.pptx", width = 170 / 25.4, height = 142 / 25.4)
```

# Appendix: Robustness checks

## A. Paired-organ intersection

```{r robustness-paired-organs, eval=FALSE}
has_above <- complete.cases(tda[, above_traits])
has_below <- complete.cases(tda[, fineroot_traits])
common_ids <- intersect(tda$ind_ID[has_above], tda$ind_ID[has_below])

tda_paired <- tda[tda$ind_ID %in% common_ids, ]
above.tda_paired <- tda_paired[, colnames(tda_paired) %in% c(
  "Comp", "Richness", "CRichGroup", "Group", "mean_phydist", "pot_number",
  "sp", "ind_ID", "pot_ID", "sp_group", above_traits)]
below.tda_paired <- tda_paired[, colnames(tda_paired) %in% c(
  "Comp", "Richness", "CRichGroup", "Group", "mean_phydist", "pot_number",
  "sp", "ind_ID", "pot_ID", "sp_group", fineroot_traits)]

# Re-run Figs 1-2/Table S1 on above.tda_paired / below.tda_paired
```

## B. Pot-level subsampling

```{r robustness-pot-subsampling, eval=FALSE}
Sample_CV_between_pot <- function(sub_df, trait_col, pot_col = "pot_ID", N = 20, iterations = 999) {
  sub_df <- sub_df[!is.na(sub_df[[trait_col]]), ]
  if (nrow(sub_df) == 0) return(NA)
  pot_ids <- sub_df[[pot_col]]
  traits <- sub_df[[trait_col]]
  unique_pots <- unique(pot_ids)
  if (length(unique_pots) <= 1) return(NA)   # need >1 pot to estimate between-pot variation

  pot_indices <- split(seq_along(traits), pot_ids)
  sample_size <- min(length(unique_pots), N)

  cv_results <- replicate(iterations, {
    sampled_idx <- vapply(pot_indices, function(idx) if (length(idx) == 1) idx else sample(idx, 1), numeric(1))
    if (length(sampled_idx) > sample_size) sampled_idx <- sample(sampled_idx, sample_size)
    CV4(traits[sampled_idx])
  })
  mean(cv_results, na.rm = TRUE)
}

# Substitute for Sample_CV() in Fig. 2 to reproduce this check
```

# Session info

```{r session-info}
sessionInfo()
```
